Expert-In-The-Loop Causal Discovery: Iterative Model Refinement Using Expert Knowledge

### 2025-02-05 paper/main.tex
Updates

In the previous section, we discussed how different types of variables require
different conditional measures of association. However, no unified measure
exists for mixed data. In this section, we introduce a novel measure of
conditional association for mixed data by extending the idea of partial
correlation coefficient to mixed data. Specifically, our approach combines a
mixed data residualization method \citep{Ankan2023} with Pillai's Trace
\citep{Pillai1955}, a multivariate measure of association based on canonical
correlations.

@@

Given a dataset $ D = (x, y, \bm{z}) $ on variables $ X $, $ Y $, and $ \bm{Z}
$, we aim to estimate the conditional association $ \rho(X, Y; \bm{Z}) $
between $ X $ and $ Y $ conditioned on $ \bm{Z} $. We first use a mixed data
residualization approach \citep{Ankan2023} to compute residuals $R_X$ and $R_Y$
for $ X $ and $ Y $. The method for computing these residuals depend on the 
type of variable:

@@

We repeat the above steps for both the variables $ X $ and  $ Y $ to get
residuals $ R_X $ and $ R_Y $. Here, we have the option to choose the
estimators based on the properties (distribution, type of relationship, and so
on) of the data. Non-parametric ensemble estimators such as Random Forest and
XGBoost can handle any combination of data types while performing reasonably
well on a wide range of distributions and relationship types. For linear
relationships, we can also choose simpler models such as linear regression.

After the above residualization steps are repeated for both $ X $ and $ Y $, we
end up with two residual matrices: $ R_X $ and $ R_Y $. The type of the
variable determines the shape for these matrices. If the variable is continuous
or ordinal, we get a residual matrix of shape $ ( n \times 1 ) $ and if the
variable is categorical, we have a residual matrix of shape $ ( x \times (k-1))
$ where one of the dummy encoded variables is dropped to avoid
multicollinearity.

We now have two continuous-valued residual matrices. The next step is
to quantify the association between these residuals. For this purpose
we use canonical correlations that have been used to measure the association
between two sets of random variables.

>>>>

In the previous section, we discussed how different types of variables require
different conditional measures of association. However, no unified measure
exists for mixed data. Here, we introduce a measure of conditional association
for mixed data by extending the concept of partial correlation coefficient for
continuous variables to mixed data. Specifically, our approach combines a mixed
data residualization method \citep{Ankan2023} with Pillai's Trace
\citep{Pillai1955}, a multivariate measure of association based on canonical
correlations.

@@

Given a dataset $ D = (x, y, \bm{z}) $ on variables $ X $, $ Y $, and $ \bm{Z}
$, we goal to estimate the conditional association $ \rho(X, Y; \bm{Z}) $ . We
achieve this by first computing the residuals $ R_X $ and $ R_Y $ for $ X $ 
and $ Y $ respectively, using the following residualization approach:

@@

Here, we have the option to choose the estimators based on the properties
(distribution, type of relationship, and so on) of the data. Non-parametric
ensemble estimators such as Random Forest and XGBoost can handle any
combination of data types while performing reasonably well on a wide range of
distributions and relationship types. For linear relationships, we can also
choose simpler models such as linear regression.

Repeating the above residualization method for both $ X $ and  $ Y $, we obtain
residual matrices $ R_X $ and $ R_Y $. The type of the variable determines the
shape for these matrices. If the variable is continuous or ordinal, the
residual matrix is of shape $ ( n \times 1 ) $ and if the variable is
categorical, we have a residual matrix of shape $ ( x \times (k-1)) $ where $ k
$ is the number of categories of the variable.

The next step is to quantify the association between these residual matrices.
For this purpose we use canonical correlations that have been widely used to
measure the association between sets of random variables.
